### 2022-03-16 outputs/paper/paper.Rmd
add 2021 us gss data, data prep and work on writeup

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Roy Chan
  - Another author
thanks: "Code and data are available at: LINK."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

# Introduction

# Data
The survey data that is presented within this paper is obtained from the 2021 United States General Social Survey (GSS).

## Survey methodology
The US GSS was created to collect public opinion and sentiment within the United States. The main component of the GSS is the replicating core, a central set of questions repeated in each year for the purpose of exploring trends in opinion over time. Additional modules present but not always consistently asked from year to year include topical questions relevant to the year. A major strength of the US GSS is in its long existence and replicating core, allowing tracking of contemporary sentiments over a significant period of time. Prior to 2021, surveys were administered in person resulting

Historically, the GSS has been administered in person but owing to the COVID-19 pandemic, the 2021 edition

# Results

# Discussion

## First discussion point

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Supplementary survey

\newpage

# References

### 2022-03-16 scripts/01-data_cleaning.R

@@ -1,35 +1,68 @@
 #### Preamble ####
-# Purpose: Clean the survey data downloaded from [...UPDATE ME!!!!!]
-# Author: Rohan Alexander [CHANGE THIS TO YOUR NAME!!!!]
-# Data: 3 January 2021
-# Contact: rohan.alexander@utoronto.ca [PROBABLY CHANGE THIS ALSO!!!!]
+# Purpose: Organise and prepare the 2021 US GSS data. 
+# Author: Roy Chan
+# Data: 20 March 2022
+# Contact: rk.chan@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: 
-# - Need to have downloaded the ACS data and saved it to inputs/data
-# - Don't forget to gitignore it!
-# - Change these to yours
-# Any other information needed?
+# - Need to download the 2021 survey data from NORC and save to inputs/data
+# - Download links can be found in the README
 
 
 #### Workspace setup ####
-# Use R Projects, not setwd().
 library(haven)
 library(tidyverse)
-# Read in the raw data. 
-raw_data <- readr::read_csv("inputs/data/raw_data.csv"
-                     )
-# Just keep some variables that may be of interest (change 
-# this depending on your interests)
-names(raw_data)
 
-reduced_data <- 
-  raw_data %>% 
-  select(first_col, 
-         second_col)
-rm(raw_data)
-         
+# Read in the data from 2021 GSS. Formatted as .dta, so use haven 
+raw_data <- haven::read_dta("inputs/data/2021_stata/gss2021.dta")
 
-#### What's next? ####
+# Select variables of interest. Variable information from GSS Codebook 
+selected_data <- raw_data %>% select(age, finalter, finrela, goodlife, satfin)
+
+# Reduce individual ages into grouped categories
+selected_data <- selected_data %>% mutate(age = case_when(
+  age >= 18 & age <= 29 ~ "18-29",
+  age >= 30 & age <= 39 ~ "30-39",
+  age >= 40 & age <= 49 ~ "40-49",
+  age >= 50 & age <= 64 ~ "50-64",
+  age >= 64 & age <= 89 ~ "64-89"
+))
+
+# Replace response values with corresponding label for each variable
+
+selected_data <- selected_data %>% mutate(finalter = case_when(
+  finalter == 1 ~ "Getting better",
+  finalter == 2 ~ "Getting worse",
[…]

### 2022-03-17 outputs/paper/paper.Rmd
paper continued

@@
 ---
-title: "My title"
-subtitle: "My subtitle if needed"
+title: "Effects of the COVID-19 pandemic on economic well-being in the United States"
+subtitle: ""
 author: 
   - Roy Chan
-  - Another author
 thanks: "Code and data are available at: LINK."
@@
 knitr::opts_chunk$set(echo = TRUE)
-
 library(tidyverse)
@@
 ## Survey methodology
-The US GSS was created to collect public opinion and sentiment within the United States.
-The main component of the GSS is the replicating core, a central set of questions repeated in each year for the purpose of exploring trends in opinion over time.
-Additional modules present but not always consistently asked from year to year include topical questions relevant to the year.
+The US GSS was created to collect long term public opinion and sentiment within the United States.
+The main component of the survey is the replicating core, a central set of questions repeated in each year the survey was conducted for the purpose of exploring trends in opinion over time (CITATION US GSS 2021).
+Additional modules present but not always consistently asked from year to year include topics relevant for the year.
+Examples of this for 2021 included but not limited to: attitudes towards policing, racial discrimination, and attitudes relating to trust in institutions and the government dealing with COVID-19.
 A major strength of the US GSS is in its long existence and replicating core, allowing tracking of contemporary sentiments over a significant period of time.
-Prior to 2021, surveys were administered in person resulting
+Prior to 2021, surveys were administered in person in an interview format by an interviewer, which as noted by the 2021 codebook resulted in high response rates of above 50%.
 
-Historically, the GSS has been administered in person but owing to the COVID-19 pandemic, the 2021 edition
+The target population of the GSS are adults over the age of 18 who lived in non-institutional housing at the time of taking the survey.
+The intention is to make the obtained sample an accurate representation of the national population of the United States, thus a stratified sampling method was chosen.
+Since 2004, the US GSS has used the sampling frame adapted from the list of postal addresses maintained by the United States Postal Service (USPS).
+Areas are categorised with respect to their population densities as either urban or rural.
+
+However, a significant limitation as discussed in the GSS Cumulative Codebook (CITATION CUMULATIVE CODEBOOK) it is more often the case that urban and rural areas are interspersed.
+This causes difficulty for interviewers to effectively pursue interviewees, due to differences in density and ability to locate households in rural areas.
+From this, a multi-stage cluster sampling approach is applied.
+Firstly, stratification is performed- the entire sampling frame is grouped consisting of urban areas and one or more of its surrounding rural areas from which about half is selected.
+Secondly, each chosen area is divided into two segments according to whether they are part of a census tract or a block group.
+This allows for generating a high quality list of clustered addresses for interviewers to efficiently conduct the survey, reducing cost.
+
+As stated by the GSS Cumulative Codebook, as of 2004 non-response of the survey has been addressed through sub-sampling.
+For all of the households in the sample that were non-respondents, a proportion, typically 50%, of those who were labelled as temporary non-respondents were pursued again.
+As such, those responses were weighted at 2 for the sub-sample to represent the entirety of the temporary non-respondents.
+
+### Methodology changes to 2021 GSS
+Owing to the COVID-19 pandemic, the 2021 edition was primarily conducted through the web instead, with an alternative phone interview option.
+Participants were recruited through mail, who were then invited to complete the survey online.
+Given that this method was utilised, the consideration of an interviewer having to physically perform interviews was no longer required.
+Thus, a stratified non-clustered sampling approach was used for a greater sample size.
+
+Another significant change was that treatment non-respondents was not handled
 
 # Results
+The survey data, along with the corresponding codebook, was retrieved from the GSS web portal (CITATION GSS WEBSITE).
+All analysis was performed in `R` [@citeR], with the data being read and prepared using `haven` [@citeHaven] and `dplyr` [@citeDplyr] respectively.
+Graphs were generated using `ggplot2` [@citeGgplot].
+
+For the purpose of this paper, we will be focusing on the sub-section of economic well-being within the replicating core, specifically in the 2018 and 2021 US GSS.
 
@@
 ## First discussion point
-
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
 
@@
 
-Weaknesses and next steps should also be included.
+- difference in survey methodology
+- survey format in person vs 
+- 2021 codebook stating higher than normal non-response for factual information
 

### 2022-03-18 outputs/paper/paper.Rmd
add 2018 data, more writeup and update gitignore

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+# hold figures in place solution (header_fix.tex) from  https://stackoverflow.com/a/33801326
+knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "!H")
+library(here)
 library(tidyverse)
@@
 
-Another significant change was that treatment non-respondents was not handled
+Another significant change was that non-respondents was not handled through sub-sampling.
 
@@
 
-For the purpose of this paper, we will be focusing on the sub-section of economic well-being within the replicating core, specifically in the 2018 and 2021 US GSS.
+For the purpose of this paper, we will be focusing on the sub-section of economic well-being within the replicating core, specifically in the 2018 and 2021 rounds of the GSS.
+Of interest are the following questions:
+
+Changes in financial situation - "During the last few years, has your financial situation been getting better, worse, or has it stayed the same?"
+
+Opinion on family income - "Compared with American families in general, would you say your family income is far below average, below average, average, above average, or far above average?"
+
+Standard of living will improve - "The way things are in America, people like me and my family have a good chance of improving our standard of living -- do you agree or disagree?"
+
+Satisfaction with financial situation - "We are interested in how people are getting along financially these days.
+So far as you and your family are concerned, would you say that you are pretty well satisfied with your present financial situation, more or less satisfied, or not satisfied at all?"
 
 # Discussion
+```{r}
+gss_data <- read_csv(here::here("inputs/data/gss_data.csv"))
+```
 
-## First discussion point
+## Changes in financial situation
 
-## Second discussion point
+```{r finalter, fig.align="center", fig.height = 6, fig.width = 5}
+# get proportion that each response represents, for corresponding age group and year
+finalter <- gss_data %>% 
+  drop_na() %>% 
+  count(age, year, finalter) %>%
+  group_by(age, year) %>%
+  mutate(pct = (n / sum(n)) * 100)
 
-## Third discussion point
+# set levels of variable, so that it is ordered from best -> worst in facet_wrap
+finalter <- finalter %>% 
+  mutate(finalter = factor(finalter, levels = c('Getting better', 'Stayed the same', 'Getting worse')))
 
-## Weaknesses and next steps
+# plot the percentages, facet_wrap to separate plots by opinion
+finalter %>% ggplot(aes(x = year, y = pct, colour = age)) + 
+  geom_point() + 
+  geom_line() +
+  scale_x_discrete(limits=c(2018, 2021)) +
+  facet_wrap(~finalter, scales = 'free_y', ncol = 1) +
+  labs(title = 'Opinions of financial situation', x = 'Year', y = 'Percentage of group with opinion (%)', color = 'Age group') + 
+  theme(text = element_text(size = 10), plot.title = element_text(face = 'bold')) +
+  theme_linedraw() + 
+  scale_colour_brewer(palette = "Set1")
+```
+- 18-29 age group saw the most significant change in opinion, from being the group who least considered their financial situation to be getting worse to the most.
+- All groups had saw a decrease in the response of 'stayed the same', indicating some change
+- 40-49, 64-89 saw an increase in 'getting better'; 50-64 almost no change between years (38.69% in 2018, 38.93% in 2021)
+
+## Opinion on family income
+```{r finrela}
+# get proportion that each response represents, for corresponding age group and year
+finrela <- gss_data %>% 
+  drop_na() %>% 
+  count(age, year, finrela) %>%
+  group_by(age, year) %>%
+  mutate(pct = (n / sum(n)) * 100)
+
+# set levels of variable, so that it is ordered from best -> worst in facet_wrap
+finrela <- finrela %>% 
+  mutate(finrela = factor(finrela, levels = c('Far above average', 
+                                              'Above average', 
+                                              'Average',
+                                              'Below average',
+                                              'Far below average')))
+
+# plot the percentages, facet_wrap to separate plots by opinion
+finrela %>% ggplot(aes(x = year, y = pct, colour = age)) + 
+  geom_point() + 
+  geom_line() +
+  scale_x_discrete(limits=c(2018, 2021)) +
+  facet_wrap(~finrela, scales = 'free_y', ncol = 1) +
+  labs(title = '', x = 'Year', y = 'Percentage of group with opinion (%)', color = 'Age group') + 
+  theme(text = element_text(size = 10), plot.title = element_text(face = 'bold')) +
+  theme_linedraw() + 
+  scale_colour_brewer(palette = "Set1")
+```
+- note that this is not objective, reliant on subject opinion
+- all age groups became more opinionated in 2021- decrease in 'average' response
+- larger portion of all groups consider income to be far above average and above average in 2021
+- below average- increases for 18-29 and 64-89
+- far below average - increases in 18-29, 30-39, 40-49
+
+## Standard of living will improve
+```{r goodlife}
+# get proportion that each response represents, for corresponding age group and year
+goodlife <- gss_data %>% 
+  drop_na() %>% 
+  count(age, year, goodlife) %>%
+  group_by(age, year) %>%
+  mutate(pct = (n / sum(n)) * 100)
+
+# set levels of variable, so that it is ordered from best -> worst in facet_wrap
+goodlife <- goodlife %>% 
+  mutate(goodlife = factor(goodlife, levels = c('Strongly agree', 
+                                              'Agree', 
+                                              'Neither agree nor disagree',
+                                              'Disagree',
+                                              'Strongly disagree')))
+
+# plot the percentages, facet_wrap to separate plots by opinion
+goodlife %>% ggplot(aes(x = year, y = pct, colour = age)) + 
+  geom_point() + 
+  geom_line() +
+  scale_x_discrete(limits=c(2018, 2021)) +
+  facet_wrap(~goodlife, scales = 'free_y', ncol = 1) +
+  labs(title = '', x = 'Year', y = 'Percentage of group with opinion (%)', color = 'Age group') + 
+  theme(text = element_text(size = 10), plot.title = element_text(face = 'bold')) +
+  theme_linedraw() + 
+  scale_colour_brewer(palette = "Set1")
+```
+
+## Satisfaction with financial situation
+```{r satfin}
+# get proportion that each response represents, for corresponding age group and year
+satfin <- gss_data %>% 
+  drop_na() %>% 
+  count(age, year, satfin) %>%
+  group_by(age, year) %>%
+  mutate(pct = (n / sum(n)) * 100)
+
+# set levels of variable, so that it is ordered from best -> worst in facet_wrap
+satfin <- satfin %>% 
+  mutate(satfin = factor(satfin, levels = c('Pretty well satisfied', 'More or less satisfied', 'Not satisfied')))
+
+# plot the percentages, facet_wrap to separate plots by opinion
+satfin %>% ggplot(aes(x = year, y = pct, colour = age)) + 
+  geom_point() + 
+  geom_line() +
+  scale_x_discrete(limits=c(2018, 2021)) +
+  facet_wrap(~satfin, scales = 'free_y', ncol = 1) +
+  labs(title = '', x = 'Year', y = 'Percentage of group with opinion (%)', color = 'Age group') + 
+  theme(text = element_text(size = 10), plot.title = element_text(face = 'bold')) +
+  theme_linedraw() + 
+  scale_colour_brewer(palette = "Set1")
+```
+- more divisive and mixed opinions than prior questions
+
+## Weaknesses
 
 - difference in survey methodology
-- survey format in person vs 
-- 2021 codebook stating higher than normal non-response for factual information
+- survey format in person vs web/phone
+- question wording for financial changes: "last few years", recency bias?
+- 2021 codebook stating higher than normal non-response for factual information, response bias
+
+## Next steps
+
+- under representation of certain populations in 2021 GSS, using weights to adjust
+- missing questions from the replicating core in 2021, specifically regarding job security
 

### 2022-03-18 scripts/01-data_cleaning.R

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Organise and prepare the 2021 US GSS data. 
+# Purpose: Organise and prepare the 2021 & 2018 US GSS data. 
 # Author: Roy Chan
 # Data: 20 March 2022
 # Contact: rk.chan@mail.utoronto.ca
@@ -14,13 +14,13 @@ library(haven)
 library(tidyverse)
 
 # Read in the data from 2021 GSS. Formatted as .dta, so use haven 
-raw_data <- haven::read_dta("inputs/data/2021_stata/gss2021.dta")
+raw_data_2021 <- haven::read_dta("inputs/data/2021_stata/gss2021.dta")
 
 # Select variables of interest. Variable information from GSS Codebook 
-selected_data <- raw_data %>% select(age, finalter, finrela, goodlife, satfin)
+selected_data_2021 <- raw_data_2021 %>% select(age, finalter, finrela, goodlife, satfin)
 
 # Reduce individual ages into grouped categories
-selected_data <- selected_data %>% mutate(age = case_when(
+selected_data_2021 <- selected_data_2021 %>% mutate(age = case_when(
   age >= 18 & age <= 29 ~ "18-29",
   age >= 30 & age <= 39 ~ "30-39",
   age >= 40 & age <= 49 ~ "40-49",
@@ -30,7 +30,7 @@ selected_data <- selected_data %>% mutate(age = case_when(
 
 # Replace response values with corresponding label for each variable
 
-selected_data <- selected_data %>% mutate(finalter = case_when(
+selected_data_2021 <- selected_data_2021 %>% mutate(finalter = case_when(
   finalter == 1 ~ "Getting better",
   finalter == 2 ~ "Getting worse",
   finalter == 3 ~ "Stayed the same"
@@ -38,7 +38,7 @@ selected_data <- selected_data %>% mutate(finalter = case_when(
 
 # FINRELA 5 was labelled as "far about average"- assumed typo as it is named
 # "far above average" in past years
-selected_data <- selected_data %>% mutate(finrela = case_when(
+selected_data_2021 <- selected_data_2021 %>% mutate(finrela = case_when(
   finrela == 1 ~ "Far below average",
   finrela == 2 ~ "Below average",
   finrela == 3 ~ "Average",
[…]

### 2022-03-18 outputs/paper/paper.Rmd
add citations, more writeup

--- 2022-03-17 outputs/paper/paper.Rmd
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+# hold figures in place solution (header_fix.tex) from  https://stackoverflow.com/a/33801326
+knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
+library(here)
 library(tidyverse)
@@
 # Introduction
+This paper was produced using `knitr` [@citeKnitr] and `bookdown` [@citeBookdown].
 
 # Data
-The survey data that is presented within this paper is obtained from the 2021 United States General Social Survey (GSS).
+The survey data that is presented within this paper is obtained from the 2018 and 2021 rounds of United States General Social Survey (US GSS).
+The data, along with its corresponding codebooks, was retrieved from the GSS web portal [@citePortal].
+All analysis was performed in `R` [@citeR], with the data being read and prepared using `haven` [@citeHaven] and `dplyr` [@citeDplyr] respectively.
+Graphs were generated using `ggplot2` [@citeGgplot].
 
-## Survey methodology
 The US GSS was created to collect long term public opinion and sentiment within the United States.
-The main component of the survey is the replicating core, a central set of questions repeated in each year the survey was conducted for the purpose of exploring trends in opinion over time (CITATION US GSS 2021).
+The main component of the survey is the replicating core, a central set of questions repeated in each year the survey was conducted for the purpose of exploring trends in opinion over time [@citeCC2021].
 Additional modules present but not always consistently asked from year to year include topics relevant for the year.
-Examples of this for 2021 included but not limited to: attitudes towards policing, racial discrimination, and attitudes relating to trust in institutions and the government dealing with COVID-19.
+Examples of this for 2021 included but were not limited to: attitudes towards policing, racial discrimination, and opinions relating to trust in institutions and the government for handling the COVID-19 pandemic.
 A major strength of the US GSS is in its long existence and replicating core, allowing tracking of contemporary sentiments over a significant period of time.
@@
 
+For the purpose of this paper, we will be focusing on the sub-section of economic well-being within the replicating core. Of interest are the following topics and corresponding questions:
+
+Changes in financial situation - "During the last few years, has your financial situation been getting better, worse, or has it stayed the same?"
+
+Opinion on family income - "Compared with American families in general, would you say your family income is far below average, below average, average, above average, or far above average?"
+
+Standard of living will improve - "The way things are in America, people like me and my family have a good chance of improving our standard of living -- do you agree or disagree?"
+
+Satisfaction with financial situation - "We are interested in how people are getting along financially these days.
+So far as you and your family are concerned, would you say that you are pretty well satisfied with your present financial situation, more or less satisfied, or not satisfied at all?"
+
+## Survey methodology
 The target population of the GSS are adults over the age of 18 who lived in non-institutional housing at the time of taking the survey.
 The intention is to make the obtained sample an accurate representation of the national population of the United States, thus a stratified sampling method was chosen.
-Since 2004, the US GSS has used the sampling frame adapted from the list of postal addresses maintained by the United States Postal Service (USPS).
+Since 2004, the US GSS has used the sampling frame adapted from the list of postal addresses maintained by the United States Postal Service.
 Areas are categorised with respect to their population densities as either urban or rural.
 
-However, a significant limitation as discussed in the GSS Cumulative Codebook (CITATION CUMULATIVE CODEBOOK) it is more often the case that urban and rural areas are interspersed.
+However, a significant limitation as discussed in the cumulative codebook [@citeCC] it is more often the case that urban and rural areas are interspersed.
 This causes difficulty for interviewers to effectively pursue interviewees, due to differences in density and ability to locate households in rural areas.
@@
 
-As stated by the GSS Cumulative Codebook, as of 2004 non-response of the survey has been addressed through sub-sampling.
-For all of the households in the sample that were non-respondents, a proportion, typically 50%, of those who were labelled as temporary non-respondents were pursued again.
+As of 2004, non-response of the survey has been addressed through sub-sampling.
+For all of the households in the sample that were non-respondents a proportion, typically 50%, of those who were labelled as temporary non-respondents were pursued again.
 As such, those responses were weighted at 2 for the sub-sample to represent the entirety of the temporary non-respondents.
@@
 ### Methodology changes to 2021 GSS
-Owing to the COVID-19 pandemic, the 2021 edition was primarily conducted through the web instead, with an alternative phone interview option.
-Participants were recruited through mail, who were then invited to complete the survey online.
+Owing to the COVID-19 pandemic, it was not possible to retain the physical interview format as used in all other years.
+Thus, the 2021 edition was primarily conducted through the web instead, with an alternative phone interview option.
+Participants were recruited through mail, which invited them to complete the survey online.
 Given that this method was utilised, the consideration of an interviewer having to physically perform interviews was no longer required.
 Thus, a stratified non-clustered sampling approach was used for a greater sample size.
-
-Another significant change was that treatment non-respondents was not handled
+Another significant change was that non-respondents was not handled through sub-sampling.
+Instead, as specified by the 2021 codebook, adjustment in response weights in accordance with known population totals for each category the response belonged to was performed iteratively.
 
 # Results
-The survey data, along with the corresponding codebook, was retrieved from the GSS web portal (CITATION GSS WEBSITE).
-All analysis was performed in `R` [@citeR], with the data being read and prepared using `haven` [@citeHaven] and `dplyr` [@citeDplyr] respectively.
-Graphs were generated using `ggplot2` [@citeGgplot].
+Of the four chosen topics which we have focused on within this paper, all have displayed some level of shift in opinion between the years of 2018 and 2021.
+Generally, it appears economic well-being has decreased in 2021 overall, with fewer positive sentiments being recorded in most topics.
 
-For the purpose of this paper, we will be focusing on the sub-section of economic well-being within the replicating core, specifically in the 2018 and 2021 US GSS.
+For the topic of changes in financial situation (Figure \@ref(fig:finalter)), all groups had saw a decrease in the response of 'Stayed the same', and reflected in increases for both positive and negative responses.
+The 18-29 age group saw the most significant change in any of the responses.
+Of all age groups, in 2018 they had the fewest percentage which considered their financial situation to be getting worse, to becoming the most in 2021.
+For ages 40-49 and 64-89, they saw an increase in the opinion of 'Getting better'; 50-64s almost no change between years (38.69% for 2018, 38.93% for 2021).
+
+Perceptions of relative family income (Figure \@ref(fig:finrela)) also saw numerous changes in responses.
+It should be noted that as a result of the question phrasing, it is heavily reliant on subject perception on what they considered as the average income for an American family.
+Nonetheless, it is shown that all groups became more opinionated in 2021 by the decrease in 'average' responses.
+A greater portion of every group considered their income to be far above and above average in 2021, and only the groups 18-29 and 64-89 had increases in 'below average', while 'far below average' responses increased for 18-29, 30-39 and 40-49.
+
+With regards to the belief that standards of living will improve under the current circumstances in the United States (Figure \@ref(fig:goodlife)), previously all groups maintained a positive sentiment in 2018 which was not reflected in 2021.
+All groups had reduced agreement, which shifted to neutral to strongly disagreeing.
+
+Lastly, satisfaction with personal and family income (Figure \@ref(fig:satfin)) became more neutral for all groups except 30-39, 40-49.
+These groups saw slight increases (<5%) in positive responses- however it was coupled alongside greater increase in negative ones.
+Similarly to changes in financial situations, 18-29 saw the largest decrease in positive satisfaction (-9.72%), with some becoming neutral (+2.25%) and more negative (+7.24%).
+
+```{r import_data}
+gss_data <- read_csv(here::here("inputs/data/gss_data.csv"))
+```
+
+```{r finalter, fig.align="center", fig.height = 6, fig.width = 5, fig.cap = 'Percentage of age group sharing opinion of changes in financial situation'}
+# get proportion that each response represents, for corresponding age group and year
+finalter <- gss_data %>% 
+  drop_na() %>% 
+  count(age, year, finalter) %>%
+  group_by(age, year) %>%
+  mutate(pct = (n / sum(n)) * 100)
+
+# set levels of variable, so that it is ordered from best -> worst in facet_wrap
+finalter <- finalter %>% 
+  mutate(finalter = factor(finalter, levels = c('Getting better', 'Stayed the same', 'Getting worse')))
+
+# plot the percentages, facet_wrap to separate plots by opinion
+finalter %>% ggplot(aes(x = year, y = pct, colour = age)) + 
+  geom_point() + 
+  geom_line() +
+  scale_x_continuous(breaks = c(2018, 2021)) +
+  facet_wrap(~finalter, scales = 'free_y', ncol = 1) +
+  labs(title = 'Opinions of changes in financial situation', x = 'Year', y = '(%)', color = 'Age group') + 
+  theme(text = element_text(size = 10), plot.title = element_text(face = 'bold')) +
+  theme_linedraw() + 
+  scale_colour_brewer(palette = "Set1")
+```
+
+```{r finrela, fig.align="center", fig.height = 7, fig.width = 5, fig.cap = 'Percentage of age group sharing perception of relative family income'}
+# get proportion that each response represents, for corresponding age group and year
+finrela <- gss_data %>% 
+  drop_na() %>% 
+  count(age, year, finrela) %>%
+  group_by(age, year) %>%
+  mutate(pct = (n / sum(n)) * 100)
+
+# set levels of variable, so that it is ordered from best -> worst in facet_wrap
+finrela <- finrela %>% 
+  mutate(finrela = factor(finrela, levels = c('Far above average', 
+                                              'Above average', 
+                                              'Average',
+                                              'Below average',
+                                              'Far below average')))
+
+# plot the percentages, facet_wrap to separate plots by opinion
+finrela %>% ggplot(aes(x = year, y = pct, colour = age)) + 
+  geom_point() + 
+  geom_line() +
+  scale_x_continuous(breaks = c(2018, 2021)) +
+  facet_wrap(~finrela, scales = 'free_y', ncol = 1) +
+  labs(title = 'Perception of family income relative to others', x = 'Year', y = '(%)', color = 'Age group') + 
+  theme(text = element_text(size = 10), plot.title = element_text(face = 'bold')) +
+  theme_linedraw() + 
+  scale_colour_brewer(palette = "Set1")
+```
+
+```{r goodlife, fig.align="center", fig.height = 7, fig.width = 5,fig.cap = 'Percentage of age group sharing belief that standard of living is likely to improve'}
+# get proportion that each response represents, for corresponding age group and year
+goodlife <- gss_data %>% 
+  drop_na() %>% 
+  count(age, year, goodlife) %>%
+  group_by(age, year) %>%
+  mutate(pct = (n / sum(n)) * 100)
+
+# set levels of variable, so that it is ordered from best -> worst in facet_wrap
+goodlife <- goodlife %>% 
+  mutate(goodlife = factor(goodlife, levels = c('Strongly agree', 
+                                              'Agree', 
+                                              'Neither agree nor disagree',
+                                              'Disagree',
+                                              'Strongly disagree')))
+
+# plot the percentages, facet_wrap to separate plots by opinion
+goodlife %>% ggplot(aes(x = year, y = pct, colour = age)) + 
+  geom_point() + 
+  geom_line() +
+  scale_x_continuous(breaks = c(2018, 2021)) +
+  facet_wrap(~goodlife, scales = 'free_y', ncol = 1) +
+  labs(title = 'Standard of living is likely to improve...', subtitle = '...given current circumstances in the U.S.', x = 'Year', y = '(%)', color = 'Age group') + 
+  theme(text = element_text(size = 10), plot.title = element_text(face = 'bold')) +
+  theme_linedraw() + 
+  scale_colour_brewer(palette = "Set1")
+```
+
+```{r satfin, fig.align="center", fig.height = 6, fig.width = 5, fig.cap = 'Percentage of age group sharing opinion of satisfaction with financial situation'}
+# get proportion that each response represents, for corresponding age group and year
+satfin <- gss_data %>% 
+  drop_na() %>% 
+  count(age, year, satfin) %>%
+  group_by(age, year) %>%
+  mutate(pct = (n / sum(n)) * 100)
+
+# set levels of variable, so that it is ordered from best -> worst in facet_wrap
+satfin <- satfin %>% 
+  mutate(satfin = factor(satfin, levels = c('Pretty well satisfied', 'More or less satisfied', 'Not satisfied')))
+
+# plot the percentages, facet_wrap to separate plots by opinion
+satfin %>% ggplot(aes(x = year, y = pct, colour = age)) + 
+  geom_point() + 
+  geom_line() +
+  scale_x_continuous(breaks = c(2018, 2021)) +
+  facet_wrap(~satfin, scales = 'free_y', ncol = 1) +
+  labs(title = "Satisfaction of personal and family's financial situation", x = 'Year', y = '(%)', color = 'Age group') + 
+  theme(text = element_text(size = 10), plot.title = element_text(face = 'bold')) +
+  theme_linedraw() + 
+  scale_colour_brewer(palette = "Set1")
+```
 
@@
 
-## First discussion point
+## Possible increase of income expectation for those aged 18-29
+- conflict between income satisfaction and above average income relative to others
 
-## Second discussion point
+## Unanimous decreased belief of improving standard of living
 
-## Third discussion point
+## Weaknesses
 
-## Weaknesses and next steps
+- should have leveraged biggest benefit US GSS and analysed more years
+- difference in survey methodology
+- survey format in person vs web/phone
+- question wording for financial changes: "last few years", recency bias?
+- 2021 codebook stating higher than normal non-response for factual information, response bias
 
-- difference in survey methodology
-- survey format in person vs 
-- 2021 codebook stating higher than normal non-response for factual information
+## Next steps
+
+- under representation of certain populations in 2021 GSS, using weights to adjust
+- missing questions from the replicating core in 2021, specifically regarding job security
 

### 2022-03-19 outputs/paper/paper.Rmd
add introduction, add more references

@@
 ---
-title: "Effects of the COVID-19 pandemic on economic well-being in the United States"
+title: "Increased financial dissatisfaction and worsening economic well-being in the United States (2018-2021)"
 subtitle: ""
@@
 ```{r setup, include=FALSE}
-# hold figures in place solution (header_fix.tex) from  https://stackoverflow.com/a/33801326
 knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
@@
 # Introduction
+Throughout the past few years, we have seen numerous major economic events that have greatly influenced economic well-being.
+Supply constraints have created shortages for many consumer goods, driving up demand, increasing cost of living and thereby raising rates of inflation.
+Stock market crashes and industry shut downs lead to high levels of unemployment during most of 2020 which is still recovering to this day.
+Many if not all of these events can be attributed in part or fully to the COVID-19 pandemic, which has majorly disrupted global circumstances.
+Within the United States, there has been a growing movement commonly known as the 'Great Resignation', characterized by high levels of resignations and voluntary unemployment of workers.
+Worker dissatisfaction has been growing within the U.S.
+in part due to salary stagnation and perceived income inequality [@citePalladino], with the pandemic acting as a catalyst for the birth of the movement.
+
+Since 1972, the United States General Social Survey (US GSS) has been collecting data on opinions within the national population of America.
+Notably, economic well-being has been a consistent topic of interest within the survey, and as such we would like to explore whether the sentiments on topics such as satisfaction of financial situation and standard of living has changed in the most recent round of the survey, as compared to the one prior.
+
 This paper was produced using `knitr` [@citeKnitr] and `bookdown` [@citeBookdown].
@@
 # Data
-The survey data that is presented within this paper is obtained from the 2018 and 2021 rounds of United States General Social Survey (US GSS).
+The survey data that is presented within this paper is obtained from the 2018 and 2021 rounds of the US GSS.
 The data, along with its corresponding codebooks, was retrieved from the GSS web portal [@citePortal].
-All analysis was performed in `R` [@citeR], with the data being read and prepared using `haven` [@citeHaven] and `dplyr` [@citeDplyr] respectively.
+All analysis was performed in `R` [@citeR], with the data being read and prepared using `haven` [@citeHaven], `here` [@citeHere], and `dplyr` [@citeDplyr].
 Graphs were generated using `ggplot2` [@citeGgplot].
@@
   facet_wrap(~finalter, scales = 'free_y', ncol = 1) +
-  labs(title = 'Opinions of changes in financial situation', x = 'Year', y = '(%)', color = 'Age group') + 
+  labs(title = 'Changes in financial situation over past few years', x = 'Year', y = '(%)', color = 'Age group') + 
   theme(text = element_text(size = 10), plot.title = element_text(face = 'bold')) +
@@
 ```
+\newpage
 
@@
 
-## Possible increase of income expectation for those aged 18-29
-- conflict between income satisfaction and above average income relative to others
+## Growing work dissatisfaction and possible increase of income expectation for those aged 18-29
+There is an interesting observation that can be made for respondents aged 18-29 that despite perceiving their financial situation as worsening in 2021, reinforced by a similar decrease in satisfaction, there has still been an increase in those who believe their income as being above average.
+We hypothesize that this seemingly contradictory combination of opinions could be the result of increasing income expectations and overall job dissatisfaction.
+Within the period of when the 2021 GSS had been conducted was the growing trend of the "Great Resignation", an economic movement of consisting of large numbers of voluntary resignations.
+In the United States, this was evidenced by worker quit rates rising drastically [@citeJOLTS] beginning in the middle of 2021, and has remained consistently high since.
+The trend has been theorized as being caused by a number of different factors, many of which have been catalyzed by the COVID-19 pandemic.
+Changes in mindsets and priorities such as establishing a healthier work-life balance and preference for remote work are common reasons for quitting [@citeHsu].
+Of all workers, the trend appears to be most prevalent in younger ages [@citeSmart], which aligns with the opinions we have seen within the US GSS.
 
 ## Unanimous decreased belief of improving standard of living
+In a similar sense, re-evaluating priorities in life can also be the reason for the decreased agreement for an improving standard of living during 2021- it could be the case that the standard of living has increased due to the shift in attitudes away from a work-centric lifestyle.
+What is notable about this topic in particular is in its unanimous change in opinion, indicating that situational changes in the U.S.
+has been felt across all age groups.
+Although not specified, it is probable that pandemic related changes are the most pronounced reason for this, causing the most significant life alterations to the national population.
+At the time of writing this paper many of the previous pandemic-related restrictions have been lifted, but at the same time major world events have occurred; it would be valuable for this topic to be explored again in the next round of the GSS to see whether this opinion holds or improves.
+
+## Perception of income inequality
+Opinions of relative financial situation in 2021 differed significantly from that of 2018, with no age group having more than 43% considering their income to be average, as compared to the lowest in 2018 of 40%.
+The decreased prominence of the average response has consequently increased that of both above and below average, possibly demonstrating a larger divide in income equality.
+The data for this topic is particularly intriguing as it also reflects perceived changes in income for "American families in general".
+Once again, this can be correlated with the events of the pandemic, disproportionately affecting those with higher living costs such as large families [@citeUnicef] and is also dependent on employment sector, ranging from little to no changes in working conditions to furloughs or permanent layoffs.
 
 ## Weaknesses
-
-- should have leveraged biggest benefit US GSS and analysed more years
-- difference in survey methodology
-- survey format in person vs web/phone
-- question wording for financial changes: "last few years", recency bias?
-- 2021 codebook stating higher than normal non-response for factual information, response bias
+Several weaknesses of this paper include the utilisation of only two rounds of the US GSS in our analysis- it would have been beneficial to leverage this significant advantage of the survey and explore trends over a greater period of time.
+For the findings of change in opinion between 2018 and 2021, it is important to note the difference in survey methodology could have influenced responses.
+Higher than normal non-responses were recorded, indicating methodology changes has altered data collection.
+Furthermore, there is a level of uncertainty due to the subjective wording in questions.
+For instance, changes in financial situation uses the wording "last few years".
+Particularly for our discussion around the effects of the pandemic, this makes it difficult to say with certainty whether it caused changes in opinion.
+On the other hand, cognitive biases are often present in assessing financial situations [@citeBansal], so there is potential for recency bias due to the weight and scale of changes due to the pandemic.
 
 ## Next steps
-
-- under representation of certain populations in 2021 GSS, using weights to adjust
-- missing questions from the replicating core in 2021, specifically regarding job security
+The 2021 GSS has differed in its respondent demographic as compared to prior years, potentially due to format changes resulting in technical limitations.
+It would be appropriate to weight responses to correct sampling bias, and/or devising a way to address non-responses.
+It is unfortunate that due to limits in survey length some questions in prior years' replicating core that would have been valuable to our analysis were not reproduced, namely job security and ease of finding a similar job.
+As such, we aim to address this by including the questions in our supplemental survey.
 

### 2022-03-20 outputs/paper/paper.Rmd
add supplemental survey link and image to appendix
add abstract

@@
 title: "Increased financial dissatisfaction and worsening economic well-being in the United States (2018-2021)"
-subtitle: ""
 author: 
@@
 date: "`r format(Sys.time(), '%d %B %Y')`"
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "Multiple notable economic events have occurred both globally and within the United States have occurred between the time of the last two rounds of the US GSS in 2018 and 2021.
+Within the GSS we investigate topics relating to economic well-being during those periods, specifically changes and satisfaction with personal and family financial situations, perception of income and standard of living.
+We find a trend of worse economic well-being, decreased belief in improvement of standard of living, and growing financial dissatisfaction especially among younger adults.
+We hypothesize that this aligns with the increase in popularity of work reform movements in the United States.
+To establish a greater correlation, we devise a supplemental survey for the GSS exploring sentiments towards work."
 output:
@@
 
-Changes in financial situation - "During the last few years, has your financial situation been getting better, worse, or has it stayed the same?"
+\textbf{Changes in financial situation} 
+"During the last few years, has your financial situation been getting better, worse, or has it stayed the same?"
 
-Opinion on family income - "Compared with American families in general, would you say your family income is far below average, below average, average, above average, or far above average?"
+\textbf{Opinion on family income} 
+"Compared with American families in general, would you say your family income is far below average, below average, average, above average, or far above average?"
 
-Standard of living will improve - "The way things are in America, people like me and my family have a good chance of improving our standard of living -- do you agree or disagree?"
+\textbf{Standard of living will improve} 
+"The way things are in America, people like me and my family have a good chance of improving our standard of living -- do you agree or disagree?"
 
-Satisfaction with financial situation - "We are interested in how people are getting along financially these days.
+\textbf{Satisfaction with financial situation}
+"We are interested in how people are getting along financially these days.
 So far as you and your family are concerned, would you say that you are pretty well satisfied with your present financial situation, more or less satisfied, or not satisfied at all?"
@@
 # Supplementary survey
+Link to the supplemental survey: https://forms.gle/c8oybFBxwQmfraUk6
+
+For reference/backup, below is a screenshot of the survey questions as they would be presented to a respondent.
+
+\includegraphics[width=15cm]{survey.jpg}
 

### 2022-03-20 outputs/paper/paper.Rmd
update readme, change name of script file, fix typo in abstract

@@
   - Roy Chan
-thanks: "Code and data are available at: LINK."
+thanks: "Code and data are available at: https://github.com/chan-roy/EcoWellBeingUS"
 date: "`r format(Sys.time(), '%d %B %Y')`"
-abstract: "Multiple notable economic events have occurred both globally and within the United States have occurred between the time of the last two rounds of the US GSS in 2018 and 2021.
-Within the GSS we investigate topics relating to economic well-being during those periods, specifically changes and satisfaction with personal and family financial situations, perception of income and standard of living.
+abstract: "Multiple notable economic events have occurred both globally and within the United States within the last few years.
+We take a look at the two most recent rounds of the US GSS in 2018 and 2021 and investigate topics relating to economic well-being during those periods, specifically changes and satisfaction with personal and family financial situations, perception of income, and standard of living.
 We find a trend of worse economic well-being, decreased belief in improvement of standard of living, and growing financial dissatisfaction especially among younger adults.

### 2022-03-20 scripts/01-data_prep.R

@@ -16,7 +16,7 @@ library(tidyverse)
 # Read in the data from 2021 GSS. Formatted as .dta, so use haven 
 raw_data_2021 <- haven::read_dta("inputs/data/2021_stata/gss2021.dta")
 
-# Select variables of interest. Variable information from GSS Codebook 
+# Select variables of interest. Variable information from GSS Codebook 2021
 selected_data_2021 <- raw_data_2021 %>% select(age, finalter, finrela, goodlife, satfin)
 
 # Reduce individual ages into grouped categories
@@ -65,7 +65,7 @@ selected_data_2021['year'] = 2021
 # Read in the data from 2018 GSS. Formatted as .dta, so use haven 
 raw_data_2018 <- haven::read_dta("inputs/data/GSS2018.dta")
 
-# Select variables of interest
+# Select variables of interest same as 2021
 selected_data_2018 <- raw_data_2018 %>% select(age, finalter, finrela, goodlife, satfin)
 
 # Reduce individual ages into grouped categories
